Hippocampal segmentation with MAGeT

### 2012-11-21 paper.Rnw
Updated figures with sensible labels in preparation for Mallar's talk

\documentclass{article}
\usepackage[margin=1.5in]{geometry}
\usepackage{algorithm, color,algpseudocode, amsmath, url, geometry, ctable }

\newcommand{\todo}[1]{
    \addcontentsline{tdo}{todo}{\protect{#1}}
    \textcolor{red}{#1}
}

\begin{document}
\SweaveOpts{echo = FALSE, message = FALSE, warning = FALSE, fig.align = 'center', tidy = FALSE, comment = NA, cache = TRUE}

<<loadpackages, results="hide", echo=FALSE, cache=FALSE>>=
library(ggplot2)
library(Hmisc)
library(reshape2)
library(ADNIMERGE)
@

@@

@@

The hippocampus is of particular interest to many researchers because it is
implicated in forms of brain dysfunction such as Alzheimer's disease and
schizophrenia, and has functional significance in cognitive processes such as
learning and memory.  For many research questions involving neuroimaging data,
accurate identification of the hippocampus and its subregions in participant MR
images is a necessary first step in order to analyse neuroanatomical
differences and changes in  sample populations.  \todo{expand}

Currently, the gold standard for neuroanatomical segmentation is manual
labelling by an expert human rater.  This is problematic for several reasons.
Practically, manual segmentation takes a significant investment of time and
expertise (Hammers2003) which may not be readily available to researchers or
clinicians.  <blah> estimates that expert segmentation can take anywhere
between 45 minutes to several days per image depending on the image resolution
and protocol.  The amount of data produced in neuroimaging experiments
increasingly exceeds the capacity for identification of specific
neuroanatomical structures by an expert manual rater.  It may be difficult for
laboratories to find or train, someone with the expertise to do such
segmentation.

Compounding this is that the hippocampus is a problematic structure for
segmentation. Its very definition is disputed and varies widely in the
literature (Geuze2004).  This has led to a variety of competing segmentation
protocols in the literature which researchers can adhere to, including
necessary ad-hoc modifications to adapt the definition in order to answer a
particular research question.  Efforts are underway to create a unified
protocol (Jack2011) for volumetry applications. But, even with a unified
protocol, in order to investigate certain research questions it may be
necessary to select from different hippocampal definitions. For example,
(Poppenk2011) found that overall hippocampal volume change did not predict
recollection memory performance, but dividing the hippocampus into anterior and
posterior regions did.

\todo{this bit needs to be a little clearer - was it just the division of the
HC that led to the improved understanding or was it that spatially localized
differences in volumes are a better way to describe the difference).}

\todo{- Also, other situations: developmental studies -- need different
definitions depending on the age range of subjects -- additionally, may want to
tweak definition to test certain hypotheses (e.g. include subiculum: yes/no,
more or less conservative WM/GM boundaries, compare with definitions from a
different study,, etc...).}

One can also expect neuroanatomical variability in this structure as it loses
volume through aging and in the time course of different diseases.  \todo{This
bit isn't really "joined to the previous statement"  may need a little glue
here}

Automated segmentation techniques are available and can minimize several of the
confounds often observed in studies that involve manual volumetry, such as the
time and expertise required for segmentation and intra- and inter-rater
reliability.  A class of segmentation approaches, known as atlas-based
segmentation (also, template-based segmentation), make use of expertly labeled
neuroanatomical atlases. An atlas MR image is warped to fit the subject
neuroanatomy using nonlinear registration techniques [4, 11], and then the
atlas labels are propagated via the registration transformation from the atlas
to match the unique neuroanatomy of a particular subject.  Segmentation
accuracy is affected by nonlinear registration and resampling errors, and by
the extent to which the atlas neuroanatomy differs from the underlying
neuroanatomical variability of the population under study (citation needed).
To address these limitations, several atlases can be used (multi-atlas
segmentation). Labels from each atlas are propagated to the subjects, by way of
an nonlinear image registration, producing as many labellings for each subject
as atlases. A label fusion technique is employed to merge the candidate
labellings for a subject into a single segmentation using a decision rule (for
instance, voxel-wise voting). Not all of the atlases need be used to segment
each subject. Selecting only those atlases that are most similar to the subject
image has been shown to significantly improve overall registration and
segmentation accuracy (Aljabar 2008, Collins2010, Leung2010, Lotjonen2010).

\todo{briefly discuss other automated segmentation methods, and justify our
focus on atlas-based methods}

In summary, whilst manual segmentation is the gold standard for neuroanatomical
segmentation, it is time-consuming and requires an investment in expertise and
training, and may be inflexible in the face of differing hippocampal
definitions or changes. Atlas-based automated segmentation dramatically reduces
the time needed for segmentation, and can be completed without any expert rater
as long as an appropriate atlas library is available <be more specific (e.g.
shares protocol, demographics, phenotypes of subject population)>.  In the
situation where such an atlas library does not exist, then an expert rater is
needed. Many automated approaches rely on an atlas library containing between
30 and 80 expertly labeled brains (Heckemann2006, Collins2010) which, even
though producing such a library may require less manual work than segmenting
the entire set of subjects, it still poses a substantial time commitment.

In this paper we extend our automated method for subcortical segmentation
(striatum, globus pallidus, and thalamus; Chakravarty2012) to hippocampal
segmentation. This method, known as MAGeT brain, is suited for situations where
manual effort is at a premium, and flexibility is needed in atlas definitions.
Our strategy is to bootstrap the familiar multi-atlas-based segmentation
process from only a very small number of atlases.  In our previous work we used
only a single atlas, based on serial histological slices, and validated the
algorithm on 20 adolescent MR images.  In this work, we extensively investigate
MAGeT used with differently sized atlas and template libraries, registration
algorithms, label fusion strategies, and hippocampal definitions.  We evaluate
our algorithm on a subset of images from the ADNI dataset, and compare our
segmentations to the manual segmentations provided by ADNI.  Additionally, we
evaluate the segmentations produced with MAGeT using hippocampal definitions
from five high-resolution T1 in-vivo atlases (Winterburn2012). <conclude with
summary that this is detailed validation the algorithm in order to pick the
best operating parameters, whereas our previous work showed the feasibility of
the approach>

The essential insight of MAGeT brain, generating a template library, is not
new. Heckemann2006 compared a multi-atlas-based segmentation method to that of
generating a template library from a single atlas and found poor performance so
did not investigate this approach any further. The LEAP algorithm (Wolz2010)
begins with 30 atlases and labels the most similar unlabeled images. The
labeled images from this step are included in the atlas set for the following
step, and this proceeds until all images have been labeled. <expand; include
other multi-atlas methods too? >

In this work we conduct a thorough validation of the MAGeT brain approach. Our
aim is not to improve on segmentation accuracy beyond existing methods, but
instead to provide a method that trades off manual segmentation expertise for
computational processing time whilst providing useful accuracy for the research
and clinical setting. <blah blah.. this sucks>

\section{Materials and Methods}
explain two registration methods, definitions, fusion methods. 
compare to each other, and then compare atlas-based, true multi-atlas, naive methods.
trying to make choices as to the best reg/fusion/

@@

\subsection{Segmentation Algorithm}
The segmentation approach we follow is an extension of traditional multi-atlas
segmentation. In multi-atlas segmentation, the inputs are a template library
consisting of a set of manually labelled MR images (atlases), and a set of
unlabelled MR images to be labelled.  Each unlabelled MR image is segmented by
first propagating to it the labels from the template library images via image
registration, and then these candidate labels are combined by a label fusion
method (e.g. voxel-wise majority voting). The label fusion method used may
determine which templates are used to segment a particular subject (for
instance, voxel-wise majority vote takes labels from all templates, whereas a
weighted majority vote selects only those templates most similar to the
subject).

Our extension to this approach is to add a preliminary stage in which we
construct the template library using a much smaller set of atlases and a subset
of the given unlabelled images.  To begin, we (possibly arbitrarily) select
from the unlabelled images to form the template library.  Multiple labellings
are given to each image in the template library via label propagation from each
of the atlases (label fusion is not conducted).  Traditional multi-atlas
segmentation is then used to produce segmentations for the entire set of
unlabelled images (including those used in the template library).  Our
algorithm is illustrated in pseudocode in Algorithm 1, and diagrammatically in
Figure 1.

- Essentially, we add an extra layer of label propagation between the set of
  atlases 
- This allows the template library to better reflect the constitution of the
  images to be labelled.

@@

The ANIMAL algorithm carries out Image registration is in two phases. In the
first, a 12-parameter linear transformation (3 translations, rotations, scales,
shears) is estimated between images using an algorithm that maximizes the
correlation between blurred MR intensities and gradient magnitude over the
whole brain [3]. In the second phase, nonlinear registration is completed using
the ANIMAL algorithm [4]: an iterative procedure that estimates a 3D
deformation field between two MR images. At first, large deformations are
estimated using blurred version of the input data. These larger deformations
are then input to subsequent steps where the fit is refined by estimating
smaller deformations on data blurred with a Gaussian kernel with a smaller
FWHM. The final transformation is a set of local translations defined on a bed
of equally spaced nodes that were estimated through the optimization of the
correlation coefficient. 
\todo{should add the MNI website here; also we should add the parameters we used somewhere}

@@

\begin{figure}[h]
<<ADNI-baseline-volumes-boxplot,echo=F,fig=T, fig.align='center',dependson='ADNI-baseline-volumes-prep',cache=T>>=
qplot(DX,value,data=melt(totals,id.vars=c("ID","DX"),measure.vars=c("FSL","MAPER","MAGeT","SNT"),na.rm=T),colour=variable,geom="boxplot")
@
  \caption{Comparison of HC volumes by FreeSurfer (FSF), MAGeT brain (MAGeT), MAPER, and manual (SNT).}
  \label{ADNI-baseline-volumes-boxplot}
\end{figure}

\begin{figure}[h]
<<ADNI-baseline-volumes-plot,echo=F,fig=T, fig.align='center',dependson='ADNI-baseline-volumes',cache=T>>=
qplot(SNT,value,data=melt(totals,id.vars=c("ID","SNT"),measure.vars=c("FSL","MAPER","MAGeT")),colour=variable, geom="smooth", method="lm")
@
  \caption{Comparison of HC volumes by FreeSurfer (FSF), MAGeT brain (MAGeT), MAPER, and manual (SNT).}
  \label{ADNI-baseline-volumes-plot}
\end{figure}

@@

\begin{figure}[h]
<<SZ_volumes,echo=F,fig=T, fig.align='center',dependson='ADNI-baseline-volumes',cache=T>>=
SZ_volumes=read.csv("data/SZ_volumes_by_method.csv")
qplot(Manual,MAGeT,data=SZ_volumes) + geom_smooth(method="lm")
@
  \caption{Comparison of total HC volumes for MAGeT against manually rated volumes}
  \label{SZ_volumes}
\end{figure}
\end{document}

>>>>

\SweaveOpts{echo = FALSE, message = FALSE, warning = FALSE, fig.align = 'center', tidy = FALSE, comment = NA, cache = TRUE}
<<loadpackages, results="hide", echo=FALSE, cache=FALSE>>=
library(ggplot2)
library(Hmisc)
library(reshape2)
library(ADNIMERGE)
@

\documentclass{article}
\usepackage[margin=1.5in]{geometry}
\usepackage{algorithm, color, algpseudocode, amsmath, url, geometry, ctable }
\usepackage[square,numbers]{natbib}
\setlength{\marginparwidth}{1.2in}
\newcommand{\marginnote}[1]{\-\marginpar[\raggedleft\footnotesize #1]{\raggedright\footnotesize #1}}
\newcommand{\todo}[1]{\marginnote{\textcolor{red}{TODO #1}}}
\newcommand{\mc}[1]{\todo{(MC) #1}}

%{\marginpar{ \textcolor{red}{\\ TODO: #1 \\} }
\begin{document}

@@

\begin{abstract}
Neuroimaging research often relies on automated anatomical segmentations of MR
images of the brain. Current multi-atlas based approaches provide accurate
segmentations of brain images by propagating manually derived segmentations of
specific neuroanatomical structures to unlabelled data. These approaches often
rely on a large number of such manually segmented atlases that take significant
time and expertise to produce. We present an algorithm for the automatic
segmentation of the hippocampus that minimizes the number of atlases needed
while still achieving similar accuracy to multi-atlas approaches.
\mc{Finish this... }
\end{abstract}

@@

The hippocampus is of particular interest to many researchers because it is
implicated in forms of brain dysfunction such as Alzheimer's disease and
schizophrenia, and has functional significance in cognitive processes such as
learning and memory.  For many research questions involving magnetic resonance
imaging (MRI) data accurate identification of the hippocampus and its
subregions is a necessary first step to better understand the individual
neuroanatomy of subjects.

Currently, the gold standard for neuroanatomical segmentation is manual
delineation by an expert human rater.  This is problematic for segmentation of
the hippocampus for several reasons.  First, manual segmentation takes a
significant investment of time and expertise \citep{Hammers2003} which may not
be readily available to researchers or clinicians.  Second, the amount of data
produced in neuroimaging experiments increasingly exceeds the capacity for
identification of specific neuroanatomical structures by an expert manual
rater.  Third, the true definition of hippocampal anatomy in MR images is
disputed \citep{Geuze2004}, as evidenced by efforts to create an unified
segmentation protocol \citep{Jack2011}.

Compounding each of these problems is the significant neuroanatomical
variability in the hippocampus throughout the course of aging, development, and
neuropsychiatric disorders \citep{Mouiha2011}.  Additionally, it may be
necessary to use several different hippocampal definitions or, in fact, make
specific modifications in the course of research. For example, Poppenk et al.
\citep{Poppenk2011} found that subdividing the hippocampus into anterior and
posterior regions resulted in a predictive relationship between volume
difference of those regions and recollection memory performance.  Thus, while
manual segmentation of the hippocampus is a necessary technique, to researchers
or clinicians who do not have access to the needed human expertise its use may
be infeasible.

Automated segmentation techniques overcome the need for human expertise by
performing segmentations computationally.  A popular class of automated
methods, {\it multi-atlas-based segmentation}, rely on a set of expertly
labeled neuroanatomical atlases. Each atlas is warped to fit a subject's
neuroanatomy using nonlinear registration
techniques \citep{Collins1995,Klein2009}.  Atlas labels are then transformed by
this warping and a {\it label fusion} technique, such as voxel-wise voting, is
used to merge the competing label definitions from each atlas into a final
segmentation for a subject.

Many descriptions of multi-atlas-based segmentation algorithms report relying
on an atlas library containing anywhere between 30 and 80 expertly labeled
brains \citep{Heckemann2011,Collins2010,Aljabar2009,Leung2010,Lotjonen2010}.  As
noted, the production of an atlas library requires significant manual effort,
and is limited since the choice of atlases or segmentation protocol may not
reflect the underlying neuroanatomical variability of the population under
study or be suited to answer the research questions at hand.

In this paper we propose an automated segmentation method to address the above
issues of existing multi-atlas-based methods. Principally, our method aims to
dramatically reduce the number of manually labelled atlases needed (under 10).
This is achieved by using the small atlas library to boot-strap a much larger
"template library", which is then used to segment the subjects in a similar
fashion to basic multi-atlas segmentation. This approach has the additional
advantage of using the unique subject population on hand to initialize the
segmentation process and improve accuracy.

The essential insight of generating a template library is not new.  Heckemann
\citep{Heckemann2006} compared generating a template library from a single atlas
to standard multi-atlas segmentation and found poor performance and so deemed
the approach as inviable.  The LEAP algorithm \citep{Wolz2010} proceeds by
iteratively segmenting the unlabelled image most similar to the atlas library
images and then incorporating the now-labelled image into the atlas library,
but requires 30 starting atlases.  The novelty of our method is to demonstrate
the possibility of producing comparable segmentation accuracy to these and
other multi-atlas-based methods while using significantly fewer manually
created atlases.

In our previous work \citep{Chakravarty2012}, we applied MAGeT brain to
segmentation of the human striatum, globus pallidus, and thalamus using a
single histologically-derived atlas. The main contribution of this paper is to
extend our approach to the human hippocampus and perform a thorough validation
over a range of atlas and template library sizes, which was not done in our
previous work.  Due to the small number of atlases required, our method can
easily accommodate different hippocampal definitions. Our aim is not to improve
on segmentation accuracy beyond existing methods, but instead to provide a
method that trades off manual segmentation expertise for computational
processing time while providing sufficient accuracy for clinical and research
applications.

\section{Materials and Methods}

explain two registration methods, definitions, fusion methods. 
compare to each other, and then compare atlas-based, true multi-atlas, naive methods.
trying to make choices as to the best reg/fusion/

@@

\subsection{Segmentation Algorithm}
\todo{Make sure we explain what MAGeT stands for}
In this paper, we use the term {\it atlas} to mean any manually segmented MR
image, and the term {\it atlas library} to mean a set of such images.  We use
the term {\it template} to refer to any MR image, and associated labelling,
used to segment another image, and the term {\it template library} to refer to
a set of such images.  An atlas library may be used as a template library but,
as we will discuss, a template library may also be composed of images with
computer generated labellings.

The segmentation approach we propose is best understood as an extension of
basic multi-atlas segmentation \citep{Collins2010}.  In multi-atlas segmentation,
an atlas library and unlabelled MR images are given as input.  Every atlas
image is nonlinearly registered to each unlabelled image, and then each atlas'
labels are propagated via the resulting transformations.  These labels are then
fused to produce a single, definitive segmentation by some label fusion method
(e.g.  voxel-wise majority vote).

Our extension adds a preliminary stage in which a template library is
constructed from input images, and used in place of an atlas library in the
standard multi-atlas-based method.  To create the template library, labels from
each atlas image are propagated to each template library image via the
transformation resulting from a non-linear registration between pair of images.
As a result, each template library image has a label from each atlas.
Basic multi-atlas segmentation is then used to produce segmentations for the
entire set of unlabelled images (including those images used in the template
library).

Label fusion is performed by cross-correlation weighted voting, a strategy
weighted towards an optimal combination of subjects from the template library
which has been previously shown to improve segmentation
accuracy \citep{Aljabar2009,Collins2010}. In this method, each template library
image is ranked in similarity to each unlabelled image by the normalized
cross-correlation of image intensities after linear registration in a region of
interest (ROI) generously encompassing the hippocampus.  Only the top ranked
template library image labels are used in a voxel-wise majority vote. The ROI
is heuristically defined as the extent of all atlas labels after linear
registration to the template, dilated by three voxels \citep{Chakravarty2012}.

@@

The ANIMAL algorithm carries out Image registration is in two phases. In the
first, a 12-parameter linear transformation (3 translations, rotations, scales,
shears) is estimated between images using an algorithm that maximizes the
correlation between blurred MR intensities and gradient magnitude over the
whole brain \citep{Collins}. In the second phase, nonlinear registration is completed using
the ANIMAL algorithm \citep{Collins1995}: an iterative procedure that estimates a 3D
deformation field between two MR images. At first, large deformations are
estimated using blurred version of the input data. These larger deformations
are then input to subsequent steps where the fit is refined by estimating
smaller deformations on data blurred with a Gaussian kernel with a smaller
FWHM. The final transformation is a set of local translations defined on a bed
of equally spaced nodes that were estimated through the optimization of the
correlation coefficient.

For the purposes of this work we used the
regularization parameters optimized in Robbins et al. \citep{Robbins2004}. It
should be noted that the MAGeT brain algorithm is not dependent on this, or
any, particular choice of registration method \citep{Chakravarty2012}.

\todo{should add the MNI website here; also we should add the parameters we used somewhere}

@@

\begin{figure}[h]
<<ADNI-baseline-volumes-boxplot,echo=F,fig=T, fig.align='center',dependson='ADNI-baseline-volumes-prep',cache=T>>=
qplot(DX,value,data=melt(totals,id.vars=c("ID","DX"),
      measure.vars=c("FSL","MAPER","MAGeT","SNT"),na.rm=T),
      colour=variable,geom="boxplot") + 
        xlab("Diagnosis") + ylab("Total Hippocampal Volume (cm3)")
@
  \caption{Comparison of HC volumes by FreeSurfer (FSF), MAGeT brain (MAGeT), MAPER, and manual (SNT).}
  \label{ADNI-baseline-volumes-boxplot}
\end{figure}

\begin{figure}[h]
<<ADNI-baseline-volumes-plot,echo=F,fig=T, fig.align='center',dependson='ADNI-baseline-volumes',cache=T>>=
qplot(SNT,value,data=melt(totals,id.vars=c("ID","SNT"),
      measure.vars=c("FSL","MAPER","MAGeT")),
      colour=variable, geom="smooth", method="lm") + 
    geom_point() +
    xlab("Total Hippocampal Volume (Manual Segmentation)") + 
    ylab("Total Hippocampal Volume (Automatic Segmentation)")
@
  \caption{{\bf ADNI Baseline cohort.} Comparison of HC volumes by FreeSurfer (FSF), MAGeT brain (MAGeT), MAPER, and manual (SNT).}
  \label{ADNI-baseline-volumes-plot}
\end{figure}

@@

\begin{figure}[h]
<<SZ_volumes,echo=F,fig=T, fig.align='center',dependson='ADNI-baseline-volumes',cache=T>>=
SZ_volumes=read.csv("data/SZ_volumes_by_method.csv")
qplot(Manual,MAGeT,data=SZ_volumes) + geom_smooth(method="lm") + 
    xlab("Total Hippocampal Volume (Manual Segmentation)") + 
    ylab("Total Hippocampal Volume (MAGeT Segmentation)")

@
  \caption{{\bf First Episode Schizophrenic Patients.} Comparison of total HC volumes for MAGeT against manually rated volumes of}
  \label{SZ_volumes}
\end{figure}
\end{document}

### 2012-11-21 paper.tex
Updated figures with sensible labels in preparation for Mallar's talk

\usepackage{alltt}
\usepackage[margin=1.5in]{geometry}
\usepackage{algorithm, color,algpseudocode, amsmath, url, geometry, ctable }

\newcommand{\todo}[1]{
    \addcontentsline{tdo}{todo}{\protect{#1}}
    \textcolor{red}{#1}
}

\begin{document}

@@

  \caption{Comparison of HC volumes by FreeSurfer (FSF), MAGeT brain (MAGeT), MAPER, and manual (SNT).}
  \label{ADNI-baseline-volumes-plot}
\end{figure}

@@

  \caption{Comparison of total HC volumes for MAGeT against manually rated volumes}
  \label{SZ_volumes}
\end{figure}
\end{document}

>>>>

\usepackage{alltt}
\usepackage[margin=1.5in]{geometry}
\usepackage{algorithm, color, algpseudocode, amsmath, url, geometry, ctable }
\usepackage[square,numbers]{natbib}
\setlength{\marginparwidth}{1.2in}
\newcommand{\marginnote}[1]{\-\marginpar[\raggedleft\footnotesize #1]{\raggedright\footnotesize #1}}
\newcommand{\todo}[1]{\marginnote{\textcolor{red}{TODO #1}}}
\newcommand{\mc}[1]{\todo{(MC) #1}}

%{\marginpar{ \textcolor{red}{\\ TODO: #1 \\} }
\begin{document}

@@

  \caption{{\em ADNI Baseline cohort} Comparison of HC volumes by FreeSurfer (FSF), MAGeT brain (MAGeT), MAPER, and manual (SNT).}
  \label{ADNI-baseline-volumes-plot}
\end{figure}

@@

  \caption{{\em First Episode Schizophrenic Patients} Comparison of total HC volumes for MAGeT against manually rated volumes of}
  \label{SZ_volumes}
\end{figure}
\end{document}
